### 2010-08-10 doc/guide/language.tex
- rearanged math terms
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3214 3e02f596-3131-0410-8352-e04eeaec37cc
- in clingo/iclingo all atoms get a default name (0) by now to make them accessible via the atomindex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3215 3e02f596-3131-0410-8352-e04eeaec37cc
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
% The input of \gringo\ and \clingo\ is the same,
% but 
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Both \gringo\ and \clingo\ can handle level-restricted input programs
(cf.\ Section~\ref{subsec:lambda}),
usually specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

In the previous sections, we have already seen a number of normal programs.
Now also considering integrity constraints, we get the following basic rule types:

\begin{tabular}{rl@{}l}
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
where $p$ is the name of some predicate, that is, a sequence of letters and digits
starting with a lowercase letter,
and any $t_i$ is a term.%
\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
  (starting with a lowercase letter).
  For such an atom~$p$, parentheses after the name are skipped, e.g.,
  \pred{p42X} could be an atom without arguments.
  Also note that lowercase letters include ``\code{\mus},''
  which can thus be used at any position within a predicate name.}
A \emph{term} $t$ starting with an uppercase letter followed by
a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
and integers are written as sequences of digits possibly preceded by ``\code{-}.''
In addition, a term can have the same syntactic structure as an atom
(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
 can be either an atom or a term, depending on where it occurs in a rule),
and functions can be nested within a term.
There are also built-in constructs
(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
having particular representations.
Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$.

In Section~\ref{subsec:semantics},
we have already provided a translation~$\psi$
from the head of a rule to a propositional formula.
By viewing an integrity constraint
$\code{:-}~\mathit{Body}\code{.}$ as a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$,
we can now explain the semantics of integrity constraints by adding the following case:
\begin{itemize}
\item $\psi[\bot]=\bot$.
\end{itemize}
Note the role of integrity constraints is to eliminate answer set candidates.
In fact, given an integrity constraint~$\code{:-}~\mathit{Body}\code{.}$,
any answer set~$X$ is such that $X\not\models\phi[\mathit{Body}]$, or
in other words, some literal~$L$ in $\mathit{Body}$ must be unsatisfied w.r.t.~$X$.
Elaborate examples on the usage of facts, rules, and integrity constraints
will be provided in Section~\ref{sec:examples}.

@@

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
A logic program~$\Pi$ containing complementary atoms is thus translated
to propositional theory
$\Phi[\Pi \cup \{\code{:-}~A\code{,}\code{-}A \mid A\in\ground{\mathcal{A}}\}]$.%
\footnote{Recall that $\ground{\mathcal{A}}$ consists of all variable-free atoms
built from predicates and functions in~$\Pi$.}
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

\begin{example}\label{ex:flies:neg}
The following logic program reformulates the one in Example~\ref{ex:flies}:
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/flycn.lp}
%
Logically, classical negation is reflected by
(implicit) integrity constraints as follows:%
\marginlabel{By invoking\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/flycn.lp}
the reader can observe that \gringo\ indeed produces the integrity constraint
in Line~7.}
%
\begin{lstlisting}[firstnumber=6]
:- flies(tux),    -flies(tux).
:- flies(tweety), -flies(tweety).
\end{lstlisting}
The additional integrity constraints do not yet change the semantics of the
program, which still has the answer sets already provided in Example~\ref{ex:flies:ground}
(of course, identifying
\code{\pred{neg\mus flies}(\const{tux})} with
\code{\pred{-flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tweety})} with
\code{\pred{-flies}(\const{tweety})}).
This situation changes if we add the following fact:
\begin{lstlisting}[firstnumber=8]
flies(tux).
\end{lstlisting}
While the program from Example~\ref{ex:flies} still admits two answer sets,
both containing 
\code{\pred{flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tux})},
there no longer is any answer set for our new program using classical negation.
In fact, answer set candidates that contain both
\code{\pred{flies}(\const{tux})} and
\code{\pred{-flies}(\const{tux})} violate the integrity constraint in Line~6.
\eexample
\end{example}

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
reflects the semantics of disjunction:
\begin{itemize}
\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
\end{itemize}
%
The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
to disjunctive programs by,
for a disjunctive rule~$r$, every predicate~$p/i$
of some atom in the head of~$r$, and each $V\in\vars{r}$,
requiring that there is some $A\in\binders{V}{r}$ such that
$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
must not contain any rule with a (non-trivial) disjunctive head.

The following rule combines the ones in Line~3 and~4 from Example~\ref{ex:flies:neg}:
\begin{lstlisting}[numbers=none]
flies(X) | -flies(X) :- bird(X).
\end{lstlisting}
In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
needs to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).

@@

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction),
\code{*} (multiplication),
\code{/} or \code{div} (integer division),
\code{mod} (modulo function),
\code{abs} (absolute value),
\code{\textasciitilde} (bitwise complement),
\code{\&}  (bitwise AND),
\code{?}   (bitwise OR), and
\code{\^} (bitwise exclusive OR).

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithf.lp}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces around tokens \code{div} and \code{mod} are mandatory.
Furthermore, the argument of function \code{abs} must be enclosed in parentheses.
In Line~9, observe that there is a unary version of \code{-},
``\code{- \var{R}}'' standing for ``\code{0 - \var{R}}.''
The four bitwise functions apply to signed integers,
using the complement on two of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not necessarily bound 
(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
only in the scope of an arithmetic function.

@@

The following built-in predicates permit term comparisons
within the bodies of rules and on the right-hand side of conditions
(cf.\ Section~\ref{subsec:gringo:condition}):
\code{==} (equal),
\code{!=} (not equal),
\code{<} (less than),
\code{<=} (less than or equal),
\code{>} (greater than),
\code{>=} (greater than or equal).

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithc.lp}}
%
\lstinputlisting{examples/arithc.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with constants,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/symbc.lp}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Finally, note that~\code{==} and~\code{!=} also apply to
compound terms over functions with non-zero arity as well as to
mixed integer and non-integer arguments.%
\footnote{%
  Such functionalities are currently not supported for the other four
  comparison predicates,
  where (after instantiation and arithmetic evaluation)
  both arguments must be either integers or (symbolic) constants.
}
\eexample
\end{example}

Importantly, a built-in comparison predicate cannot bind variables,
that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
Also note that built-in predicates are not considered by level mappings~$\xi$
in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
one may assume that built-in predicates are mapped to level~$0$ and all other
predicates to levels greater than~$0$.

@@

Built-in predicate~\code{=} can be used in the body of a rule
(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
to assign a term on its right-hand side to a variable on its left-hand side.
Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
provided that~$t$ is variable-free or that its variables are bound
by atoms other than assignments with~\var{X} on the right-hand side.
Cyclic assignments over otherwise unbound variables are thus excluded.

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

@@

\begin{example}\label{ex:int}
The next program makes use of integer intervals:
%
\lstinputlisting{examples/int.lp}
%
The facts in Line~1 and~2 are expanded as follows:
%
\begin{lstlisting}[numbers=none]
num(1).    num(2).   num(3).   num(4).   num(5).
top5(5).  top5(6).  top5(7).  top5(8).  top5(9).
\end{lstlisting}
%
By instantiating~\var{X} to~\const{9},
the rule in Line~4 becomes:
%
\begin{lstlisting}[numbers=none]
top5num(1..5,5..9) :- num(5..9), top5(1..5), top(9).
\end{lstlisting}
%
It is expanded to the cross product
$(\code{\const{1}..\const{5}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{1}..\const{5}})$
of intervals:
%
\begin{lstlisting}[numbers=none,escapechar=@]
top5num(1,5) :- num(5), top5(1), top(9).
top5num(2,5) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@
top5num(5,5) :- num(5), top5(1), top(9).
top5num(1,6) :- num(5), top5(1), top(9).
top5num(2,6) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(1), top(9).
top5num(1,5) :- num(6), top5(1), top(9).
top5num(2,5) :- num(6), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(1), top(9).
top5num(1,5) :- num(5), top5(2), top(9).
top5num(2,5) :- num(5), top5(2), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@         @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(4), top(9).
top5num(1,5) :- num(5), top5(5), top(9).
top5num(2,5) :- num(5), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(5), top(9).
top5num(1,5) :- num(6), top5(5), top(9).
top5num(2,5) :- num(6), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(5), top(9).
\end{lstlisting}
%
Note that only the rules with
\code{\pred{num}(\const{5})} and
\code{\pred{top5}(\const{5})}
\marginlabel{%
  Again the unique answer set is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/int.lp}}
%
in the body actually contribute to the unique answer set
of the above program by deriving all atoms \code{\pred{top5num}($m$,$n$)}
for $\const{1}\leq m\leq\const{5}$ and $\const{5}\leq n\leq\const{9}$.
\eexample
\end{example}

@@

\begin{example}\label{ex:cond}
The following program uses conditions in a rule body and in a rule head:
%
\lstinputlisting{examples/cond.lp}
%
We are particularly interested in the rules in Line~5 and~6,
instantiated as follows:%
%
\begin{lstlisting}[firstnumber=5]
meet :- available(jane), available(john).
on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
\end{lstlisting}
%
The conjunction in Line~5 is obtained by replacing~\var{X} in
\marginlabel{%
  The reader can reproduce these ground rules by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/cond.lp}}
\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
Furthermore, the condition in the head of the rule in Line~6 turns into
a disjunction over all ground instances of
\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
such that
\code{\pred{day}($t$)} holds.
That is, conditions in the body and in the head of a rule
are expanded to different basic language constructs.

@@

>>>>

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Usually logic programs are specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

@@

\begin{figure}
\begin{rail}
	constant    : ('\_'*) '[a-z]' ('[A-Za-z0-9]'*);
	variable    : ('\_'*) '[A-Z]' ('[A-Za-z0-9]'*);
	simpleterm  : ('integer' | constant | variable | '\_' | '\#supremum' | '\#infimum');
	function    : constant '(' simpleterm (',' simpleterm*) ')';
	term        : simpleterm | function;
\end{rail}
\caption{Terms\label{fig:terms}}
\end{figure}

Every logic program is constructed from terms.
An overview of \gringo{} terms is depicted in Figure \ref{fig:terms}.
The most basic terms are integers, constants, and variables.
Furthermore, there are some special variables and constants.
An anonymous variable denoted by `\_' is simialar to a normal variable 
but each occurrence is treated like a different variable 
(intuitively a new unique variable name is substituted).
Additionally, there are the two special constants `\#supremum' and `\#infimum'
representing the largest and the smallest possible values, respectively,
which behave essentially like constants.
Finally, there are function symbols which are composed of other terms.
A term that does not contain any (anonymous) variables is called a ground term.
More complex terms involving arithmetics and other constructs
are introduced later on.

\begin{tabular}{rl@{}l}\label{eq:normal:rule}
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
as a function symbol or constant.
Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$ where the connective \code{not} corresponds to default negation.
The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
Facts have an empty body.
Throughout this section we will further extend the predicates that can be used 
in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
Furthermore, \gringo{} expects rules to be save, i.e., 
all variables that appear in a rule have to appear in some some positive literal (a literal not preceded by \code{not}) in the body.
If a variable appears positively in some predicate, then we say that this predicate binds the variable.

Intuitively, the head of a rule has to be true whenever all its body literals are true.
In ASP every predicate needs some derivation, i.e., 
a predicate cannot be true if there is no body deriving it.
This implies that only predicates appearing in some head can appear in answer sets.
Furthermore, derivations
\footnote{There are extensions like disjunctions 
that go beyond simple derivability and also require minimality w.r.t. a reduct. 
We will not cover the semantics of such constraints in this guide.}%
have to be acyclic,
a feature that is important to model reachability.
As a simple example consider the program \code{a :- b. b :- a.}
The only answer set to this program is the empty set.
Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
Finally, note that default negation is ignored when checking for acyclic derivations 
(we do not need a reason for a predicate being false).
Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
has the two answer sets $\{a\}$ and $\{b\}$.
But in practice it is never needed to express choices this way.
For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
constraint, which provides a much more readable way to introduce choices.

A fact has an empty body and thus its associated head predicate is always true 
and will appear in all answer sets.
On the other hand, integrity constraints eliminate answer set candidates.
They are merely tests that discard unwanted answer sets.
That is, there will be no answer sets that satisfy all literals an integrity constraint.
Elaborate examples on the usage of facts, rules, and integrity constraints
will be provided in Section~\ref{sec:examples}.

@@

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

\begin{example}\label{ex:flies:neg}
Consider a logic program comprising the following facts:
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/flycn.lp}
%
Logically, classical negation is reflected by
(implicit) integrity constraints as follows:%
\marginlabel{By invoking\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}\attach{examples/bird.lp}{bird.lp}~\attach{examples/flycn.lp}{flycn.lp}}
the reader can observe that \gringo\ indeed produces the integrity constraint
in Line~7.}
%
\begin{lstlisting}[firstnumber=6]
:- flies(tux),    -flies(tux).
:- flies(tweety), -flies(tweety).
\end{lstlisting}
The program has two answer sets.
One contains \code{flies(tweety)} and the other contains \code{-flies(tweety)}.
Let us now add a new fact to the program:
\begin{lstlisting}[firstnumber=8]
flies(tux).
\end{lstlisting}
There no longer is any answer set for our new program using classical negation.
In fact, answer set candidates that contain both
\code{\pred{flies}(\const{tux})} and
\code{\pred{-flies}(\const{tux})} violate the integrity constraint in Line~6.
\eexample
\end{example}

@@

Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
A disjunction is true if at least one of its atoms is true.
Additionally, logic programs have to satisfy a minimality criterion, 
which we will not detail in this guide.
The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
does not admit the answer set ${a,b}$ because it is no minimal model.

In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
needs to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).

@@

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction, unary minus),
\code{*} (multiplication),
\code{/} or \code{\#div} (integer division),
\code{\textbackslash} or \code{\#mod} (modulo function),
\code{**} or \code{\#pow} (exponentation),
\code{|$\cdot$|} or \code{\#abs} (absolute value),
\code{\&} (bitwise AND),
\code{?} (bitwise OR),
\code{\^} (bitwise exclusive OR), and
\code{\textasciitilde} (bitwise complement).

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces after tokens \code{\#div} and \code{\#mod} are mandatory.
Furthermore, the argument of function \code{\#abs}, \code{\#div}, and \code{\#mod} must be enclosed in parentheses.
The four bitwise functions apply to signed integers,
using the complement on two of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not bound by a corresponding atom.
For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
Although, the latter might produce an infinite grounding 
and \gringo{} will not necessarily halt when given such an input.

@@

The following built-in predicates permit term comparisons
within the bodies of rules:
\code{==} (equal),
\code{!=} (not equal),
\code{<} (less than),
\code{<=} (less than or equal),
\code{>} (greater than),
\code{>=} (greater than or equal).

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
%
\lstinputlisting{examples/arithf.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with arbitrary ground terms,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Integers are compared in the usual way and constants are ordered lexicographically.
Function symbols are compared first using their arity.
If the arity differs, then the name of the function symbol is compared lexicographically.
If again the name differs, then arguments are compared component wise.
Finally, integers are always smaller than constants and constants are always smaller than function symbols.
\eexample
\end{example}

Importantly, a built-in comparison predicate cannot bind variables, i.e.,
when checking whether a rule is safe, 
comparison predicates are not considered to be positive.

@@

The built-in predicates~\code{:=} and ~\code{=} can be used in the body of a rule
to unify a term on their right-hand side 
to a (non-ground) term or variable on its left-hand side, respectively.

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/assign.lp}{assign.lp}}}%
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the right-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\begin{example}\label{ex:unify}
The second program demonstrates the usage of \code{:=},
which allows for terms on the left hand side:
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/unify.lp}{unify.lp}}}%
%
\lstinputlisting{examples/unify.lp}
%
Here the term \code{f(a,X,X+1)} is unified with every function symbol provided by \code{sym}/$1$.
Note the usage of \code{X+1} in the term.
\gringo{} will not try to unify any term containing arithmetic 
but in this example \code{X} occurs also directly as second argument of the argument
and can thus be unified with.
The term $X+1$ is merely a test that deferred and checked later.
For example, the fourth line is equivalent to:
\begin{lstlisting}[firstnumber=6]
unifyf(X) :- f(a,X,Y) := F, sym(F), Y = X + 1.
\end{lstlisting}%
\eexample
\end{example}

Finally, note that assignments to some extend can bind variables.
Of course cyclic assignments cannot bind variables.
For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
Additionally, the unification is restricted to ground terms on the right hand side of the assignment, 
that is all variables on the right hand side have to be by some other predicate.

@@

\begin{example}\label{ex:int}
The next program makes use of integer intervals:
%
\lstinputlisting{examples/int.lp}
%
The facts in Line~1 and~2 are expanded as follows:
%
\begin{lstlisting}[numbers=none]
num(1).    num(2).   num(3).   num(4).   num(5).
top5(5).  top5(6).  top5(7).  top5(8).  top5(9).
\end{lstlisting}
%
By instantiating~\var{X} to~\const{9},
the rule in Line~4 becomes:
%
\begin{lstlisting}[numbers=none]
top5num(1..5,5..9) :- num(5..9), top5(1..5), top(9).
\end{lstlisting}
%
It is expanded to the cross product
$(\code{\const{1}..\const{5}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{1}..\const{5}})$
of intervals:
%
\begin{lstlisting}[numbers=none,escapechar=@]
top5num(1,5) :- num(5), top5(1), top(9).
top5num(2,5) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@
top5num(5,5) :- num(5), top5(1), top(9).
top5num(1,6) :- num(5), top5(1), top(9).
top5num(2,6) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(1), top(9).
top5num(1,5) :- num(6), top5(1), top(9).
top5num(2,5) :- num(6), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(1), top(9).
top5num(1,5) :- num(5), top5(2), top(9).
top5num(2,5) :- num(5), top5(2), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@         @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(4), top(9).
top5num(1,5) :- num(5), top5(5), top(9).
top5num(2,5) :- num(5), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(5), top(9).
top5num(1,5) :- num(6), top5(5), top(9).
top5num(2,5) :- num(6), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(5), top(9).
\end{lstlisting}
%
Note that only the rules with
\code{\pred{num}(\const{5})} and
\code{\pred{top5}(\const{5})}
\marginlabel{%
  Again the unique answer set is obtained via call:\\
  \code{\mbox{~}gringo -t \attach{examples/int.lp}{int.lp}}}
%
in the body actually contribute to the unique answer set
of the above program by deriving all atoms \code{\pred{top5num}($m$,$n$)}
for $\const{1}\leq m\leq\const{5}$ and $\const{5}\leq n\leq\const{9}$.
\eexample
\end{example}

@@

\begin{example}\label{ex:cond}
The following program uses conditions in a rule body and in a rule head:
%
\lstinputlisting{examples/cond.lp}
%
We are particularly interested in the rules in Line~5 and~6,
instantiated as follows:%
%
\begin{lstlisting}[firstnumber=5]
meet :- available(jane), available(john).
on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
\end{lstlisting}
%
The conjunction in Line~5 is obtained by replacing~\var{X} in
\marginlabel{%
  The reader can reproduce these ground rules by invoking:\\
  \code{\mbox{~}gringo -t \attach{examples/cond.lp}{cond.lp}}}
\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
Furthermore, the condition in the head of the rule in Line~6 turns into
a disjunction over all ground instances of
\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
such that
\code{\pred{day}($t$)} holds.
That is, conditions in the body and in the head of a rule
are expanded to different basic language constructs.

@@

\com{Should we mention stratifed aggregates that do not necessarily need conditions?}
